Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-03-06 analysis/paper.qmd
Add Zenodo links
Resolve some TODOs

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [x] Discussion
  - [x] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [x] tbl-results-summary
  - [ ] tbl-predictors
  - [x] palaeodist figs
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
  - [ ] Replace "see below" etc. with section links
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

In this paper we present a complementary, deductive approach based on ecological niche modelling.
Rather than inferring environmental conditions from preserved physical evidence, 
we predict the ranges of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
<!-- TODO which we can use to ... --->
Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple past climatologies.

# Background
<!--
* The transition to agriculture in West Asia was...

@@

## Ecological niche modelling in archaeology
-->
Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

@@

<!-- TODO: rewrite -->
Occurrence data only tells us where a species is present;
there is rarely definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) random sample of points, weighted equally against the presences in the regression.
@ValaviEtAl2022 also recommend using a large background sample for random forest models.

@@

Our reconstructed palaeodistributions (shown in full in the appendix) indicate that the majority of species have experienced a significant reduction and/or shift in their ranges since the Early Holocene. 
`r n_reduced_range` of `r nrow(flora)` species have reduced ranges; `r n_reduced_range_10` of more than 10% or more.
Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further below), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
The average range of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` during the Younger Dryas (i.e. ranges recovered slightly between the Younger Dryas and Early Holocene).
This perhaps indicates that although this period is considered one of climatatic amelioration globally [@JonesEtAl2019], the colder conditions of the Pleistocene may have supported more extensive plant-based economies in West Asia specifically.

@@

Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene (@fig-palaeodist-sum).
These include *Ficus carica* (fig);
*Hordeum* spp. (wild barleys);
*Lathyrus aphaca* and *L. sativus* (both marginally edible legumes);
*Triticum aestivum compactum* (in the N. Levant), *T. monococcum aegilopoides*, *T. durum*, and *Triticum urartu* (but not the other wheat progenitor, *T. turgidum dicoccum* – see below);
*Aegilops speltoides*, but not *Aegilops tauschii* (goatgrasses);
and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
leaving the Levant and to a lesser extent the Aegean and Cyprus as refugia.

@@

## Biogeography of crop progenitors

@@

## Hindcast models do not predict archaeobotanical composition

@@

"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
<!--
* Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
* Discrepencies suggests several areas for further research and methodological development
 --> 
Modelling a large number of species using machine learning, the substantial occurrence datasets available for the present day, and a hindcasting approach to past distributions also represents a significant advance in the methodology of palaeoecological niche modelling. 
This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@KargerEtAl2017; @BrownEtAl2018].
Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
Though we have benefitted from the relatively long tradition of compiling archaeobotanical data in our region of study [@ColledgeEtAl2004; @ShennanConolly2007; @ADEMNES; @LucasFuller2018; @FullerEtAl2018; @WallaceEtAl2018; @ORIGINS], 
further development of open, comprehensive and up-to-date 'backbone' datasets on site locations and chronologies is needed to advance archaeoecological modelling to the same level.

>>>>

* [x] Stop modelling
- [x] First draft:
- [x] Figures
- [ ] Biogeography background section
- [ ] Revise discussion of archaeo verification
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Zenodo release
- [ ] Final proofread

@@

In this paper we present a complementary, deductive approach based on ecological niche modelling.
Rather than inferring environmental conditions from preserved physical evidence, 
we predict the ranges of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
This means that the ancient data can be reserved for assessing the model's ability to 'hindcast' past conditions.
In this sense, discrepencies between the two records are perhaps the most interesting result, as they indicate processes affecting one or both records that are not fully accounted for and therefore generate new questions.
Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple past climatologies.

# Background

* The transition to agriculture in West Asia was...

@@

## Ecological niche modelling in archaeology

Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

@@

<!-- TODO: rewrite -->
Random Forest is a presence–absence approach to niche modelling and therefore requires not just data on where a species is present, but where it is definitely not present.
However 'absence' data is rarely available because it requires exhaustive survey.
In practice, most applications of niche modelling are 'presence-only' and, where absence data is required (as for Random Forest), it is supplied as a random background sample of 'pseudo-absence' points.
The purpose of this sample is to inform the model about the nature of the underlying environment. 
The stochastic generation process means that some 'pseudo-absence' points will overlap or fall close to presences, so ensuring the model is not overly influenced by background samples is critical to its predictive importance [@ValaviEtAl2022].
Here we follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) uniform sample of points from across the land area of the study region.
These points are then weighted equally against the presences in the regression to produce a 'balanced Random Forest' [@ValaviEtAl2022].

@@

Our reconstructed palaeodistributions (shown in full in the appendix) indicate that the majority of species have experienced a significant reduction and/or shift in their ranges since the Early Holocene. 
`r n_reduced_range` of `r nrow(flora)` species have reduced ranges; `r n_reduced_range_10` of more than 10% or more.
Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further in @sec-discuss-hindcasting), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
The average range of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` during the Younger Dryas (i.e. ranges recovered slightly between the Younger Dryas and Early Holocene).
This perhaps indicates that although this period is considered one of climatatic amelioration globally [@JonesEtAl2019], the colder conditions of the Pleistocene may have supported more extensive plant-based economies in West Asia specifically.

@@

Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene (@fig-palaeodist-sum).
These include *Ficus carica* (fig);
*Hordeum* spp. (wild barleys);
*Lathyrus aphaca* and *L. sativus* (both marginally edible legumes);
*Triticum aestivum compactum* (in the N. Levant), *T. monococcum aegilopoides*, *T. durum*, and *Triticum urartu* (but not the other wheat progenitor, *T. turgidum dicoccum* – see @sec-discuss-crops);
*Aegilops speltoides*, but not *Aegilops tauschii* (goatgrasses);
and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
leaving the Levant and to a lesser extent the Aegean and Cyprus as refugia.

@@

## Biogeography of crop progenitors {#sec-discuss-crops}

@@

## Hindcast models do not predict archaeobotanical composition {#sec-discuss-hindcasting}

@@

"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.

* Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
* Discrepancies suggests several areas for further research and methodological development

Modelling a large number of species using machine learning, the substantial occurrence datasets available for the present day, and a hindcasting approach to past distributions also represents a significant advance in the methodology of palaeoecological niche modelling. 
This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@KargerEtAl2017; @BrownEtAl2018].
Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
Though we have benefited from the relatively long tradition of compiling archaeobotanical data in our region of study [@ColledgeEtAl2004; @ShennanConolly2007; @ADEMNES; @LucasFuller2018; @FullerEtAl2018; @WallaceEtAl2018; @ORIGINS], 
further development of open, comprehensive and up-to-date 'backbone' datasets on site locations and chronologies is needed to advance archaeoecological modelling to the same level.

# Data availability

The data and R code used to produced is archived with Zenodo at <https://doi.org/10.5281/zenodo.14629984>

### 2025-03-06 analysis/paper.tex
Add Zenodo links
Resolve some TODOs

In this paper we present a complementary, deductive approach based on
ecological niche modelling. Rather than inferring environmental
conditions from preserved physical evidence, we predict the ranges of
individual species relevant to human subsistence based on a model of
their current environmental niche and simulations of past palaeoclimate.
Though hypothetical, this gives us an independent line of evidence on
past ecologies that is independent of the environmental archaeological
and palaeoclimatic records. Our computational approach is also readily
scaled up, allowing us to model spatially-explicit palaeodistributions
for a large number of species, for the whole region, under multiple past
climatologies.

@@

@@

Occurrence data only tells us where a species is present; there is
rarely definitive information on where the species is \emph{not} found.
We therefore need to generate random background points or
``pseudo-absences'' to feed to the model. There are several ways to do
this. We follow the advice of \citet{BarbetMassinEtAl2012} for
regression-based species distribution models and use a large (≈10000)
random sample of points, weighted equally against the presences in the
regression. \citet{ValaviEtAl2022} also recommend using a large
background sample for random forest models.

@@

Our reconstructed palaeodistributions (shown in full in the appendix)
indicate that the majority of species have experienced a significant
reduction and/or shift in their ranges since the Early Holocene. 75 of
82 species have reduced ranges; 66 of more than 10\% or more. Though the
magnitude of the reduction from prehistory to the present likely
reflects a degree of overfitting in the model (discussed further below),
fluctuations in modelled range size between the Bølling-Allerød
(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
(11.7--8.3 ka) are more directly comparable
(Figure~\ref{fig-pred-range-density}). The average range of modelled
species was 16\% in the Early Holocene compared to the Bølling--Allerød,
and 25\% during the Younger Dryas (i.e.~ranges recovered slightly
between the Younger Dryas and Early Holocene). This perhaps indicates
that although this period is considered one of climatatic amelioration
globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
may have supported more extensive plant-based economies in West Asia
specifically.

@@

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene (Figure~\ref{fig-palaeodist-sum}). These
include \emph{Ficus carica} (fig); \emph{Hordeum} spp. (wild barleys);
\emph{Lathyrus aphaca} and \emph{L. sativus} (both marginally edible
legumes); \emph{Triticum aestivum compactum} (in the N. Levant),
\emph{T. monococcum aegilopoides}, \emph{T. durum}, and \emph{Triticum
urartu} (but not the other wheat progenitor, \emph{T. turgidum dicoccum}
-- see below); \emph{Aegilops speltoides}, but not \emph{Aegilops
tauschii} (goatgrasses); and \emph{Vicia} spp. (vetches), including
\emph{Vicia faba} (broad beans). most of Anatolia, Northern Mesopotamia,
and the Zagros Mountains in particular disappear from the predicted
ranges of these species, leaving the Levant and to a lesser extent the
Aegean and Cyprus as refugia.

@@

\subsection{Biogeography of crop
progenitors}\label{biogeography-of-crop-progenitors}

@@

\subsection{Hindcast models do not predict archaeobotanical
composition}\label{hindcast-models-do-not-predict-archaeobotanical-composition}

@@

``All models are wrong'' \citep{Box1976}, but the performance of our
models on independent test datasets give confidence in its predictions,
which are plausible and consistent with broad-scale patterns in the
archaeological record. Modelling a large number of species using machine
learning, the substantial occurrence datasets available for the present
day, and a hindcasting approach to past distributions also represents a
significant advance in the methodology of palaeoecological niche
modelling. This approach is enabled by the availability of high quality,
global open datasets in ecology \citep{GBIF2025, GBIFSecretariat2023},
earth science \citep{SRTM}, and climatology
\citep{KargerEtAl2017, BrownEtAl2018}. Unfortunately in relation to
these fields open data in archaeology lags conspiciously behind. Though
we have benefitted from the relatively long tradition of compiling
archaeobotanical data in our region of study
\citep{ColledgeEtAl2004, ShennanConolly2007, ADEMNES, LucasFuller2018, FullerEtAl2018, WallaceEtAl2018, ORIGINS},
further development of open, comprehensive and up-to-date `backbone'
datasets on site locations and chronologies is needed to advance
archaeoecological modelling to the same level.

>>>>

In this paper we present a complementary, deductive approach based on
ecological niche modelling. Rather than inferring environmental
conditions from preserved physical evidence, we predict the ranges of
individual species relevant to human subsistence based on a model of
their current environmental niche and simulations of past palaeoclimate.
Though hypothetical, this gives us an independent line of evidence on
past ecologies that is independent of the environmental archaeological
and palaeoclimatic records. This means that the ancient data can be
reserved for assessing the model's ability to `hindcast' past
conditions. In this sense, discrepencies between the two records are
perhaps the most interesting result, as they indicate processes
affecting one or both records that are not fully accounted for and
therefore generate new questions. Our computational approach is also
readily scaled up, allowing us to model spatially-explicit
palaeodistributions for a large number of species, for the whole region,
under multiple past climatologies.

@@

\begin{itemize}
\tightlist
\item
  The transition to agriculture in West Asia was\ldots{}
\end{itemize}

\subsection{Biogeography and agricultural
origins}\label{biogeography-and-agricultural-origins}

\begin{itemize}
\tightlist
\item
  Has always been important in study of agricultural origins

  \begin{itemize}
  \tightlist
  \item
    Historically: Vavilov, Pumpelly \& Childe
  \item
    Genetic studies tell us origin points, but not ranges
  \end{itemize}
\item
  Important to e.g.

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    \citep{MartinEtAl2016, MartinEtAl2025}
  \item
    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

  \begin{itemize}
  \tightlist
  \item
    More than the ``founder crops'';
  \item
    More than food
  \item
    (In archaebot., not all intentionally collected)
  \item
    Geographically and temporally diverse
  \item
    \ldots so we model lots of species!
  \end{itemize}
\item
  Regional ecological reconstructions generally rely on the `expert
  interpolation' (or what do they call it with isoscapes?) method

  \begin{itemize}
  \tightlist
  \item
    See CSEAS (AEA-prep) presentation
  \item
    Figure: comparisons
  \end{itemize}
\end{itemize}

\subsection{Ecological niche modelling in
archaeology}\label{ecological-niche-modelling-in-archaeology}

@@

Random Forest is a presence--absence approach to niche modelling and
therefore requires not just data on where a species is present, but
where it is definitely not present. However `absence' data is rarely
available because it requires exhaustive survey. In practice, most
applications of niche modelling are `presence-only' and, where absence
data is required (as for Random Forest), it is supplied as a random
background sample of `pseudo-absence' points. The purpose of this sample
is to inform the model about the nature of the underlying environment.
The stochastic generation process means that some `pseudo-absence'
points will overlap or fall close to presences, so ensuring the model is
not overly influenced by background samples is critical to its
predictive importance \citep{ValaviEtAl2022}. Here we follow the advice
of \citet{BarbetMassinEtAl2012} for regression-based species
distribution models and use a large (≈10000) uniform sample of points
from across the land area of the study region. These points are then
weighted equally against the presences in the regression to produce a
`balanced Random Forest' \citep{ValaviEtAl2022}.

@@

Our reconstructed palaeodistributions (shown in full in the appendix)
indicate that the majority of species have experienced a significant
reduction and/or shift in their ranges since the Early Holocene. 75 of
82 species have reduced ranges; 66 of more than 10\% or more. Though the
magnitude of the reduction from prehistory to the present likely
reflects a degree of overfitting in the model (discussed further in
Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled range
size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
(12.9--11.7 ka), and Early Holocene (11.7--8.3 ka) are more directly
comparable (Figure~\ref{fig-pred-range-density}). The average range of
modelled species was 16\% in the Early Holocene compared to the
Bølling--Allerød, and 25\% during the Younger Dryas (i.e.~ranges
recovered slightly between the Younger Dryas and Early Holocene). This
perhaps indicates that although this period is considered one of
climatatic amelioration globally \citep{JonesEtAl2019}, the colder
conditions of the Pleistocene may have supported more extensive
plant-based economies in West Asia specifically.

@@

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene (Figure~\ref{fig-palaeodist-sum}). These
include \emph{Ficus carica} (fig); \emph{Hordeum} spp. (wild barleys);
\emph{Lathyrus aphaca} and \emph{L. sativus} (both marginally edible
legumes); \emph{Triticum aestivum compactum} (in the N. Levant),
\emph{T. monococcum aegilopoides}, \emph{T. durum}, and \emph{Triticum
urartu} (but not the other wheat progenitor, \emph{T. turgidum dicoccum}
-- see Section~\ref{sec-discuss-crops}); \emph{Aegilops speltoides}, but
not \emph{Aegilops tauschii} (goatgrasses); and \emph{Vicia} spp.
(vetches), including \emph{Vicia faba} (broad beans). most of Anatolia,
Northern Mesopotamia, and the Zagros Mountains in particular disappear
from the predicted ranges of these species, leaving the Levant and to a
lesser extent the Aegean and Cyprus as refugia.

@@

\subsection{Biogeography of crop progenitors}\label{sec-discuss-crops}

@@

\subsection{Hindcast models do not predict archaeobotanical
composition}\label{sec-discuss-hindcasting}

@@

``All models are wrong'' \citep{Box1976}, but the performance of our
models on independent test datasets give confidence in its predictions,
which are plausible and consistent with broad-scale patterns in the
archaeological record.

\begin{itemize}
\tightlist
\item
  Results w.r.t. to specific archaeologically-attested species not to
  promising, but this doesn't mean the models are useless!
\item
  Discrepancies suggests several areas for further research and
  methodological development
\end{itemize}

Modelling a large number of species using machine learning, the
substantial occurrence datasets available for the present day, and a
hindcasting approach to past distributions also represents a significant
advance in the methodology of palaeoecological niche modelling. This
approach is enabled by the availability of high quality, global open
datasets in ecology \citep{GBIF2025, GBIFSecretariat2023}, earth science
\citep{SRTM}, and climatology \citep{KargerEtAl2017, BrownEtAl2018}.
Unfortunately in relation to these fields open data in archaeology lags
conspiciously behind. Though we have benefited from the relatively long
tradition of compiling archaeobotanical data in our region of study
\citep{ColledgeEtAl2004, ShennanConolly2007, ADEMNES, LucasFuller2018, FullerEtAl2018, WallaceEtAl2018, ORIGINS},
further development of open, comprehensive and up-to-date `backbone'
datasets on site locations and chronologies is needed to advance
archaeoecological modelling to the same level.

\section{Data availability}\label{data-availability}

The data and R code used to produced is archived with Zenodo at
\url{https://doi.org/10.5281/zenodo.14629984}
